Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-26 draft.tex
Revised age-gender figure and added caption

\section{The Lumbar Spine Data Set}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/age-gender}
  \caption{Age-Gender}
  \label{fig:example1}
\end{figure}
\label{sec:MaterialsAndMethod}
%%
Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

@@

\section{\uppercase{Experiments}}
\label{sec:Experiments}
\noindent Spine shape is confounded by several somatometric variables.
%%
Larger people have also a longer spine and its shape is more straight compared to smaller subjects.
%%
Since men are on average taller than women, gender is another confounder.
%%
Large body weight increases the spine load, resulting in a bend shape.
%%
All these variables need to be taken into account, when correlating spine curvature and torsion with non-image parameter.
%%
Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.
%%
To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{Klemm2014VIS}.
\subsection{Preliminary Results}
\begin{figure*}[htb]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
  \caption{%%
A generalized pairs plot of all image-derived variables \cite{GPLOMS} colored by presence or absence of back pain.
%%
Pairwise combinations of image-derived variables are visualized using scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
%%
The box plots (left) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
%%
No correlation with back pain are present can be seen in this plot.
}
  \label{fig:image-parameter-range}
\end{figure*}
%%
\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a generalized pairs plot \cite{GPLOMS}.
%%
Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
%%
Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
%%
Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
%%
The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
%%
Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
%%
The projections to the transversal planes attract attention as they have many outliers.
%%
We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
%%--------------------------------
\subsubsection{Correlation Matrix}
We calculated an association matrix to to assess correlations between the image parameters.
%%
The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal in Figure~\ref{fig:image-parameter-range}.
%%
Curvature, curvature angle and torsion correlate strongly with their planar projections.
%%
Also the mean curvature and the curvature angle correlate by a factor of -0.89.
%%
%Torsion and curvature/curvature angle do not correlate.
Torsion does not correlate with any other image-derived variable.
%%--------------------------------------------------------
\subsubsection{Correlation of Image Parameters With Back Pain}
Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters \add{as generalized pairs plot}.
%%
\rem{We created a generalized pairs plot for subject groups derived by dividing them into quantile groups of body-size.}
%%
No statistical significant correlation could be observed through all subject groups.
%%
The box plots show no difference between subjects with and without back pain.

@@

% \subsubsection{Diese Themen hier unterbringen}
% \begin{itemize}
% 	\item Direkter Vergleich
% 	\item Verteilung des Means
% 	\item Korrelationsmatrix
% 	\item PCA
% \end{itemize}
% Weil das nicht ausreichend war, haben wir einen C5.0 Baum erstellt, um alle potentiellen Zusammenhänge zwischen bildbezogenen und Nichtbildbezogenen Parametern herzustellen - Was kann man mit den Bilddaten erklären und was nicht?
%
% HetCor hier noch mit unterbringen. Damit kann man eine Korrelationsmatrix unabhängig vom Datentyp berechnen, um eine große Korrellationsmatrix zu erhalten (Supplementary Material). Hier können die Einträge in den Zeilen der Bildparameter angeschaut werden, um zu schauen, ob man eine Korrellation mit bestimmten Parametern finden kann! Das müsste auch wieder für alle Größengruppen passieren! (https://stats.stackexchange.com/questions/101016/pca-and-component-scores-based-on-a-mix-of-continuous-binary-and-categorical-va)

\subsection{Experimental Settings and Results}
Decision Tree erklaeren, ebenfalls Umrechnen in die Dummy-Variablen. Was sind die Zielvariablen
\begin{itemize}
\item Altersgruppierung
\item Male/Female
\item Gewicht (BMI - dünn, normal, dick)
\end{itemize}
% Creating a Decision Tree using C5.0

@@

A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
%%
They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~(a)).
%%
This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
%%
Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~(b).
%%
Low spine canal thickness can be an indicator for an impending herniated disk.
%%
Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
%%
The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~(c).
%% ------------------------------------
\subsection{Applicability of the Method}
%%
Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
%%
Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, making their results harder to assess and to abstract.
%%
Our method to plot derived metrics and custom tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our brushable visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

>>>>

\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

@@

\section{\uppercase{Experiments}}
\label{sec:Experiments}
\begin{figure}[!h]
  \includegraphics[width=0.5\textwidth]{figures/age-gender}
	Correlation of age and gender regarding the lumbar spine size.
	The bar chart shows subjects divided into four different age groups (x-axis) and their subject number (y-axis).
	Each bar contains the mean lumbar spine of the respective group.
	The shape color encodes the distance to the overall male (top chart) or female (bottom chart) mean shape.
	The dark gray share of each bar encodes the portion of male (top chart) or female (bottom chart) subjects.
	Women have a higher live expectance than men, hence their higher share in the old age group.
	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
	Due to bone erosion, older people are also smaller on average.
  \label{fig:age-gender}
\end{figure}
\noindent Spine shape is confounded by several somatometric variables.
Larger people have also a longer spine and its shape is more straight compared to smaller subjects.
Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also confounder (Fig.~\ref{fig:age-gender}).
Large body weight increases the spine load, resulting in a bend shape.
All these variables need to be taken into account, when correlating spine curvature and torsion with non-image parameter.
Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.
To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{Klemm2014VIS}.
\subsection{Preliminary Results}
\begin{figure*}[htb]
  \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
  \caption{%%
A generalized pairs plot of all image-derived variables \cite{GPLOMS} colored by presence or absence of back pain.
Pairwise combinations of image-derived variables are visualized using scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
The box plots (left) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
No correlation with back pain are present can be seen in this plot.
  \label{fig:image-parameter-range}
\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a generalized pairs plot \cite{GPLOMS}.
Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
The projections to the transversal planes attract attention as they have many outliers.
We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
%%--------------------------------
\subsubsection{Correlation Matrix}
We calculated an association matrix to to assess correlations between the image parameters.
The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal in Figure~\ref{fig:image-parameter-range}.
Curvature, curvature angle and torsion correlate strongly with their planar projections.
Also the mean curvature and the curvature angle correlate by a factor of -0.89.
%Torsion and curvature/curvature angle do not correlate.
Torsion does not correlate with any other image-derived variable.
%%--------------------------------------------------------
\subsubsection{Correlation of Image Parameters With Back Pain}
Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters \add{as generalized pairs plot}.
\rem{We created a generalized pairs plot for subject groups derived by dividing them into quantile groups of body-size.}
No statistical significant correlation could be observed through all subject groups.
The box plots show no difference between subjects with and without back pain.

@@

% \subsubsection{Diese Themen hier unterbringen}
% \begin{itemize}
% 	\item Direkter Vergleich
% 	\item Verteilung des Means
% 	\item Korrelationsmatrix
% 	\item PCA
% \end{itemize}
% Weil das nicht ausreichend war, haben wir einen C5.0 Baum erstellt, um alle potentiellen Zusammenhänge zwischen bildbezogenen und Nichtbildbezogenen Parametern herzustellen - Was kann man mit den Bilddaten erklären und was nicht?
%
% HetCor hier noch mit unterbringen. Damit kann man eine Korrelationsmatrix unabhängig vom Datentyp berechnen, um eine große Korrellationsmatrix zu erhalten (Supplementary Material). Hier können die Einträge in den Zeilen der Bildparameter angeschaut werden, um zu schauen, ob man eine Korrellation mit bestimmten Parametern finden kann! Das müsste auch wieder für alle Größengruppen passieren! (https://stats.stackexchange.com/questions/101016/pca-and-component-scores-based-on-a-mix-of-continuous-binary-and-categorical-va)
% Creating a Decision Tree using C5.0

@@

A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
%%
They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
%%
This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
%%
Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~b).
%%
Low spine canal thickness can be an indicator for an impending herniated disk.
%%
Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
%%
The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~c).
%% ------------------------------------
\subsection{Applicability of the Method}
%%
Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
%%
Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, making their results harder to assess and to abstract.
%%
Our method to plot derived metrics and custom tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our brushable visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%
